---
title: "Example Confirmatory Data Analysis Script"
output: html_notebook
---

## Replication study and hypotheses

This is a data analysis plan for the replication of:

Bryan, C. J., Adams, G. S., & Monin, B. (2013). When cheating would make you a cheater: implicating the self prevents unethical behavior. Journal of Experimental Psychology: General, 142(4), 1001.

The original test of the primary hypothesis is a independent t-test that compares the mean head count of reported coin flips across two randomize conditions: cheater and cheating. Power analyses revealed that 75 participants were needed in each condition to achieve 80% power with an effect size of Cohen's d = 0.46. Here I will document the analysis plan for this replication. A successful replication will find evidence of support for the alternate hypothesis, which is:

H1: The mean heads reported in the cheating condition will larger than the mean head reported in the cheater condition.  

The null hypothesis being tested in in the independent t-test is:

H0: There will be no significant difference in mean heads reported between the cheating and cheater conditions.

## Load data

```{r}
library(readr)
data <- read_csv("C:/Users/CURRANT/Dropbox/Work/LSE/PB230/MT5/Workshop/indt.csv")
```

## Preliminary analysis

### Data Screening and cleaning

First the data will be screened for erroneous and missing values. The outcome in this study, mean heads, is bounded between 0 and 10 and so any responses beyond this range will be replaced with a missing value.

```{r}
library(tidyverse)
library(psych)
library(naniar)
data <-
  data %>%
  replace_with_na(replace = list(~.heads > 10, ~.heads < 0)) 
describe(data)
```

As this is an experimental study, any missing values will be removed listwise from the sample.

```{r}
drop_na(data) 
```

### Summary statistics and distributions

Second, I will inspect the means and standard devations of each condition in the study.

```{r}
describeBy(data,data$condition)
```

Then I will create a plot for the distributions of each condition.

```{r}
library(ggplot2)
library(plyr)

## Find mean for each condition

means <- ddply(data, "condition", summarise, heads.mean=mean(heads))

## Plot distributions

plot1 <-
  ggplot(data, aes(x=heads)) + geom_histogram(binwidth=.5, colour="black", fill="white") + 
    facet_grid(condition ~ .) +
    geom_vline(data=means, aes(xintercept=heads.mean),
               linetype="dashed", size=1, colour="red") + theme_classic()

## Save plot

ggsave(file="C:/Users/CURRANT/Dropbox/Work/LSE/PB230/MT5/Workshop/plot_1.png")
plot1
```

And finally create am APA table with the mean and sd for each condition.

```{r}
library(apaTables)
apa.1way.table(iv=condition,dv=heads,data=data,filename="C:/Users/CURRANT/Dropbox/Work/LSE/PB230/MT5/Workshop/table_1.doc",table.number = 1)
```


## Confirmatory analysis

The next step is to conduct the independent samples t-test for the primary confirmatory hypothesis testing as listed in the data analysis plan. I will conduct a normal theory t-test and a bootstrapped t-test.

```{r}
library(MKinfer)
t.test <- t.test(data$heads ~ data$condition, var.equal=TRUE)
boot.t.test(data$heads ~ data$condition, var.equal=TRUE, R = 5000)
```

Now I will calculate the effect size for the experimental condition.

```{r}
library(rstatix)
cohens_d(data, heads ~ condition,  var.equal = TRUE)
```

## Tables and figures for reporting 

To report the results from the independent t-test I will visualize the means and 95% confidence intervals for each condition using a bar chart with error bars.

```{r}
library(rcompanion)
sum <- groupwiseMean(heads ~ condition, data=data,conf   = 0.95,digits = 3)
qplot(x    = condition ,
      y    = Mean,
      data = sum) +

geom_bar(stat ="identity",
         fill = "gray", width = .40, alpha=.5
        ) +

geom_errorbar(aes(ymin  = Trad.lower,
                  ymax  = Trad.upper,
                  width = 0.15)) +
  xlab("Condition") +
  ylab("Mean Number of Heads")+
  theme_classic()+
  ggsave(file="C:/Users/CURRANT/Dropbox/Work/LSE/PB230/MT5/Workshop/plot_2.png")
```
